---
title: 'Final Project: Bike Allocation'
author: "Ishita Agrawal"
date: "2024-06-03"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 1. Collecting and Preparing Data

### 1. Loading and Knowing the Data

```{r}

library(e1071)
library(dplyr)
library(ggplot2)
library(tidyr)
library(lubridate)
library(caret)
library(data.table)
library(car)

setwd("C:/Users/admin/Desktop/depaul/fundamentals/final group project")

data <- read.csv("citibikeDemand.csv")

dim(data)
colnames(data)
head(data)
tail(data)
str(data)

data$DemandDate <- as.Date(data$DemandDate, format = "%m/%d/%Y")
data$DemandTime <- as.factor(data$DemandTime)
data$DayOfWeek <- as.factor(data$DayOfWeek)
#Monday = 1 and Sunday = 7
```

### 2. Summary Statistics

```{r}

summary(data[c("TripDurationMinutes", "StartPerCapitaIncome", "EndPerCapitaIncome", "StartPctHouseholdsNoVehicle", "EndPctHouseholdsNoVehicle")])

summary(data)
```

### 4. Handeling Missing Data

```{r}

# Remove rows with missing values
data_clean <- data %>%
  mutate(across(everything(), ~ifelse(is.na(.), mean(., na.rm = TRUE), .)))

# View the cleaned data
summary(data_clean)

# Convert 'DemandDate' to Date type if it's not already
data_clean$DemandDate <- as.Date(data_clean$DemandDate)
```

## 2. Descriptive Analysis: Identify Demand Patterns

### 1. Univariate Descriptive Analysis

```{r univariate analysis}

# Mean and Standard Deviation
mean_demand <- mean(data_clean$Demand, na.rm = TRUE)
sd_demand <- sd(data_clean$Demand, na.rm = TRUE)

# Skewness and Kurtosis
skewness_demand <- skewness(data_clean$Demand, na.rm = TRUE)
kurtosis_demand <- kurtosis(data_clean$Demand, na.rm = TRUE)

cat("Mean of Demand:", mean_demand, "\n")
cat("Standard Deviation of Demand:", sd_demand, "\n")
cat("Skewness of Demand:", skewness_demand, "\n")
cat("Kurtosis of Demand:", kurtosis_demand, "\n")
```

### 2. Bivariate Analysis

```{r Bivariate Analysis}

correlation <- cor(data_clean$Demand, data_clean$Temperature, use = "complete.obs")

# Covariance
covariance <- cov(data_clean$Demand, data_clean$Temperature, use = "complete.obs")

cat("Correlation between Demand and Temperature:", correlation, "\n")
cat("Covariance between Demand and Temperature:", covariance, "\n")

# Summarize demand by day of the week
demand_summary <- data_clean %>%
  group_by(DayOfWeek) %>%
  summarize(
    MeanDemand = mean(Demand, na.rm = TRUE),
    MedianDemand = median(Demand, na.rm = TRUE),
    SD = sd(Demand, na.rm = TRUE),
    MinDemand = min(Demand, na.rm = TRUE),
    MaxDemand = max(Demand, na.rm = TRUE)
  )

# Print the summary
print(demand_summary)
```

### 3. Visualization using ggplot2

```{r Visualization ggplot2}

#1. Demand Distribution Plot
ggplot(data_clean, aes(x = Demand)) +
  geom_histogram(bins = 150, fill = "skyblue", color = "black") +
  labs(title = "Distribution of Bike Demand", x = "Demand", y = "Count")

# 2. Demand by Time of Day
demand_by_time <- data_clean %>%
  group_by(DemandTime) %>%
  summarize(avg_demand = mean(Demand, na.rm = TRUE))

ggplot(demand_by_time, aes(x = DemandTime, y = avg_demand)) +
  geom_col(fill = "skyblue") +
  labs(title = "Average Demand by Time of Day", x = "Time of Day", y = "Average Demand")

# 3. demand by day of week
ggplot(data_clean, aes(x = factor(DayOfWeek), y = Demand)) +
  geom_boxplot(fill = "skyblue") +
  labs(title = "Demand by Day of Week", x = "Day of Week", y = "Demand")

# 4. Demand vs. Temperature
ggplot(data_clean, aes(x = Temperature, y = Demand)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(title = "Demand vs. Temperature", x = "Temperature", y = "Demand")

# 5. Demand vs. Precipitation

ggplot(data_clean, aes(x = DailyPrecipitation, y = Demand)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(title = "Demand vs. Daily Precipitation", x = "Daily Precipitation", y = "Demand")

library(maps)
library(ggmap)

# Create a data frame with station locations
station_locations <- data_clean %>%
  select(StartStationLatitude, StartStationLongitude, StartStationName) %>%
  distinct() %>%
  rename(Latitude = StartStationLatitude, Longitude = StartStationLongitude, StationName = StartStationName)

# Get the map boundaries
map_bounds <- c(left = min(station_locations$Longitude), right = max(station_locations$Longitude),
                bottom = min(station_locations$Latitude), top = max(station_locations$Latitude))

# Create a base map
base_map <- get_map(location = map_bounds, source = "stadia", maptype = "toner")

# Plot station locations on the map
ggmap(base_map) +
  geom_point(data = station_locations, aes(x = Longitude, y = Latitude), color = "red", size = 2) +
  labs(title = "Station Locations")

```

### 4. Findings from Descriptive Analysis

## 3. Predictive Analysis: Demand Forecast

### 1. Logistic Regression Model

```{r Linear Regression Model}

str(data_clean)

data_clean$DemandTime <- as.factor(data_clean$DemandTime)
data_clean$StartStationId <- as.factor(data_clean$StartStationId)
data_clean$EndStationId <- as.factor(data_clean$EndStationId)

data_clean <- as.data.table(data_clean)

str(data_clean)

reg <- lm(Demand ~ DemandTime + StartStationId + EndStationId, data = data_clean)

summary(reg)


```

### 3. Predict Future Demand for Daytime and Evening

```{r}

# Filter the new data for specific stations (assuming you have selected station IDs 3183, 3202, 3184, 3185, and 3187)
selected_stations <- c(3674, 3671, 3631, 3680, 3673)


# Split the data into daytime and evening
daytime_data <- 
  data.table(
      DemandTime = 1,
      StartStationId = c(3664, 3671, 3631, 3680, 3673, 3664, 3671, 3631, 3680, 3673, 3664, 3671, 3631, 3680, 3673, 3664, 3671, 3631, 3680, 3673, 3664, 3671, 3631, 3680, 3673),
      EndStationId = c(rep(3674, 5), rep(3671, 5), rep(3631, 5), rep(3680, 5), rep(3673, 5)),
      Temperature = c(84, 84, 85, 85, 86, 86, 87, 87, 88, 88, 89, 89, 90, 90, 91, 91, 90, 80, 89, 89, 88, 87, 86, 84, 82),
      DailyPrecipitation = 0,
      DemandDate = as.Date("2018-09-06"),
      DayOfWeek = 4,
      Month = 9
  )

eveningtime_data <- 
  data.table(
      DemandTime = 2,
      StartStationId = c(3664, 3671, 3631, 3680, 3673, 3664, 3671, 3631, 3680, 3673, 3664, 3671, 3631, 3680, 3673, 3664, 3671, 3631, 3680, 3673, 3664, 3671, 3631, 3680, 3673),
      EndStationId = c(rep(3674, 5), rep(3671, 5), rep(3631, 5), rep(3680, 5), rep(3673, 5)),
      Temperature = c(rep(70, 12), rep(85, 13)),
      DailyPrecipitation = 0,
      DemandDate = as.Date("2018-09-06"),
      DayOfWeek = 4,
      Month = 9
  )

daytime_data <- daytime_data %>%
  mutate(DemandTime = factor(DemandTime),
         StartStationId = factor(StartStationId),
         EndStationId = factor(EndStationId))

eveningtime_data <- eveningtime_data %>%
  mutate(DemandTime = factor(DemandTime),
         StartStationId = factor(StartStationId),
         EndStationId = factor(EndStationId))

# Predict demand for daytime data
daytime_data$PredictedDemand <- predict(reg, newdata = daytime_data)

# Predict demand for evening data
eveningtime_data$PredictedDemand <- predict(reg, newdata = eveningtime_data)

eveningtime_data
daytime_data
```
